---
title: "COVID-19 Outcomes"
author: "Alana Schreibman"
date: "5/27/2021"
output: 
  html_document:
    toc: true
    toc_float: 
        collapsed: true
        smooth_scroll: true
    depth: 3 
    theme: cosmo
    highlight: pygments
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

***

The purpose of this file is to practice applying basic R commands learned through lecture videos and perform exploratory analysis on worldwide COVID-19 data. 


### Our World in Data GitHub COVID-19 outcomes
Data was read in and summarized as shown below: (current file is as of 5/31)
```{r, eval = TRUE}
owid <- read.csv(url("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv"), na.strings = c("", "NA"), colClasses = c("date" = "Date"), header = TRUE) 
dim(owid)
```

```{r, eval = TRUE}
head(owid)
str(owid)
summary(owid)
```

### Full Cleaned Data Set
Data set, including possible confounders for use in later analysis, is cleaned and filtered below. 
```{r, eval = TRUE}
library(dplyr)
library(lubridate)
library(knitr)
owid.filtered <- owid %>%
                  rename(iso.code = iso_code, totcases = total_cases, totdeaths = total_deaths, totcases.mil = total_cases_per_million, totdeaths.mil = total_deaths_per_million, hosp.patients = hosp_patients, hosp.patients.mil = hosp_patients_per_million, med.age = median_age, older70 = aged_70_older, gdp.cap = gdp_per_capita, card.death = cardiovasc_death_rate, diab.prev = diabetes_prevalence, fem.smokers = female_smokers, male.smokers = male_smokers) %>%
                  select(c(iso.code:totcases, totdeaths, totcases.mil, totdeaths.mil, hosp.patients, hosp.patients.mil, med.age, older70, gdp.cap, card.death, diab.prev, fem.smokers, male.smokers)) %>% #Renamed variables and filtered to keep only those of interest. 
                  filter(date == "2020-05-31") #Filtered out all but most recent date (most recent date is a running sum of total cases).
dim(owid.filtered)
head(owid.filtered)
tail(owid.filtered)
table(owid.filtered$continent, useNA = "ifany") #Data includes both continent and country-specific outcomes. Continent outcomes can be distinguished by "NA" in the continent column; the continent name is in the location column. 
```

#### Continent Analysis
Data is filtered to remove entries for specific country locations and include only data that aggregates COVID-19 outcomes on a continent and world scale. 
```{r, eval = TRUE}
owid.continents <- owid.filtered %>%
                    filter(is.na(continent)) %>%
                    select(-continent) #Used missingness properties to select only continent rows and then removed continent column. 
print(owid.continents)
table(owid.continents$location)
```

#### Country Analysis
Data is filtered to include only country-specific COVID-19 outcomes and exclude continent and world-wide data.  
```{r, eval = TRUE}
owid.countries <- owid.filtered %>%
                    filter(!is.na(continent)) #Removed all rows with NA for continent, leaving behind only countries. 
print(owid.countries)
table(owid.countries$continent)
```

#### Tests for Missingness
```{r, eval = TRUE}
table(owid.filtered$totcases, useNA = "ifany") %>%
  tail() #No missing values for total cases column. 
table(owid.filtered$totdeaths, useNA = "ifany") %>%
  tail() #Several missing values for total deaths column. Keep this in mind when analyzing. 
```

#### Modified Data Frames
Filtered OWID data frames further narrowed to COVID-19 cases only, deaths only, and hospitalizations only, separated by continent and by country respectively. 
```{r, eval = FALSE}
#By continent:
select(owid.continents, c(iso.code:totcases, totcases.mil)) #COVID-19 cases by continent. 
select(owid.continents, c(iso.code:date, totdeaths, totdeaths.mil)) #COVID-19 deaths by continent. 
select(owid.continents, c(iso.code:date, hosp.patients, hosp.patients.mil)) #COVID-19 hospitalizations by continent. 

#By country:
select(owid.countries, c(iso.code:totcases, totcases.mil)) #COVID-19 cases by country. 
select(owid.countries, c(iso.code:date, totdeaths, totdeaths.mil)) #COVID-19 deaths by country. 
select(owid.countries, c(iso.code:date, hosp.patients, hosp.patients.mil)) #COVID-19 hospitalizations by country.  
```

### Continent COVID-19 Plots
Plots for exploratory analysis comparing COVID-19 cases by continent for both total cases and total cases per million people.
```{r, eval = TRUE}
library(ggplot2)
library(cowplot)
owid.countries %>%
    group_by(continent) %>%
    summarise(totcases = mean(totcases.mil, na.rm = TRUE)) #Mean COVID-19 cases by continent.
ggplot(data = owid.continents, aes(x = location, y = totcases)) +
  geom_bar(stat="identity") +
  labs(title = "Total COVID-19 Cases by Continent") +
  labs(x = "Continent", y = "Total COVID-19 Cases") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) #Barplot of total cases by continent. 
ggplot(data = owid.continents, aes(x = location, y = totcases.mil)) +
  geom_bar(stat="identity") + 
  labs(title = "Total COVID-19 Cases per Million People by Continent") +
  labs(x = "Continent", y = "Total COVID-19 Cases per Million") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) #Barplot of total cases per million by continent. 
```

### Top 10 Country COVID-19 Plots
Barplots filtering out the 10 countries with the highest COVID-19 cases per million people on each continent are the outputs of the function below: 
```{r, eval = TRUE}
top10countries <- function(x) { 
                    owid.countries %>% 
                    filter(continent == x) %>%
                    arrange(desc(totcases.mil)) %>%
                    head(n=10) %>%
                    ggplot(aes(x = location, y = totcases.mil)) +
                        geom_bar(stat = "identity") +
                        labs(x = "Country", y = "Total COVID-19 Cases per Million People") +
                        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
                        labs(title = paste0("COVID-19 Cases per Million People in ", (continent = x)))
}
for (i in c("North America", "Europe", "Asia", "Africa", "Oceania", "South America")) {
  print(top10countries(i))
} #Loops function for continents of interest. 
plot_grid() #Use this to combine the graphs from the for loop. Need to figure out how to name the graphs first? 
```

### GDP and COVID-19 Cases per Million
Scatter plots with a smooth line showing the relationship of GDP per capita and COVID-19 cases per million people are the outputs of the function below: 
```{r, eval = TRUE}
cases.gdp <- function(x) { 
                    owid.countries %>% 
                    filter(continent == x) %>%
                    ggplot(aes(x = gdp.cap, y = totcases.mil)) +
                        geom_point(color = "#fee6ce") +
                        geom_smooth(color = "red") +
                        theme_bw() +
                        labs(x = "GDP per Capita", y = "Total COVID-19 Cases per Million People") +
                        labs(title = paste0("COVID-19 Cases per Million People in ", (continent = x), " vs. GDP per Capita"))
} 
for (i in c("North America", "Europe", "Asia", "Africa", "Oceania", "South America")) {
  print(cases.gdp(i))
} #Loops function for continents of interest. 
```

### Median Age and COVID-19 Cases per Million
Cut median age in two halves and convert to factors with labels "below.med" and "above.med". Bar plot shows COVID-19 cases per million for each country separated by continent, with fill color indicating whether the median population age is above or below the global median. 
```{r, eval = TRUE}
summarise(owid.countries, med.age = mean(med.age, na.rm = TRUE)) #Mean of the median age of all countries.
cases.med.age <- function(x) { 
                    owid.countries %>%
                    mutate(med.age = cut(med.age, breaks = c(0, 30.38, 48.3), labels = c("below.med", "above.med"))) %>%
                    mutate(med.age <- as.factor(med.age)) %>%
                    filter(continent == x) %>%
                    ggplot(aes(x = location, y = totcases.mil, fill = factor(med.age))) +
                        geom_bar(stat = "identity", position = "dodge") +
                        theme_bw() +
                        labs(x = "Country", y = "Total COVID-19 Cases per Million People") +
                        scale_fill_brewer(palette = "Set1") +
                        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
                        labs(title = paste0("COVID-19 Cases per Million People by Country in ", (continent = x), " and Median Age"))
} 
for (i in c("North America", "Europe", "Asia", "Africa", "Oceania", "South America")) {
  print(cases.med.age(i))
} #Loops function for continents of interest. 
```
### Smoking and COVID-19 Cases per Million 
Graphs below show percentages of female and male smokers by country and the correlation to COVID-19 cases. 
```{r, eval = TRUE}
cases.smoking <- function(x) {
                    owid.countries %>%
                    filter(continent == x) %>%
                    ggplot(aes(x = fem.smokers, y = totcases)) +
                      geom_point(color = "#333aff", alpha = 1) +
                      geom_smooth(color = "red") +
                      theme_bw() +
                      labs(x = "Percentage of Female Smokers in Country", y = "Total COVID-19 Cases per Million People") +
                      labs(title = paste0("COVID-19 Cases per Million People in ", (continent = x), " vs. Percentage of Female Smokers"))
}
for (i in c("North America", "Europe", "Asia", "Africa", "Oceania", "South America")) {
  print(cases.smoking(i)) #Apply function to each continent. 
}

cases.smoking <- function(x) {
                    owid.countries %>%
                    filter(continent == x) %>%
                    ggplot(aes(x = male.smokers, y = totcases)) +
                      geom_point(color = "#333aff", alpha = 1) +
                      geom_smooth(color = "red") +
                      theme_bw() +
                      labs(x = "Percentage of Male Smokers in Country", y = "Total COVID-19 Cases per Million People") +
                      labs(title = paste0("COVID-19 Cases per Million People in ", (continent = x), " vs. Percentage of Male Smokers"))
}
for (i in c("North America", "Europe", "Asia", "Africa", "Oceania", "South America")) {
  print(cases.smoking(i)) #Apply function to each continent. 
}
```



