---
title: "COVIDmortality.Rmd"
author: "Alana Schreibman"
date: "5/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


***
The purpose of this file is to practice applying basic R commands learned through lecture videos and perform exploratory analysis on worldwide COVID-19 data. 
***


###Our World in Data GitHub COVID-19 outcomes
Data was downloaded and summarized as shown below:
```{r, eval = FALSE}
owid <- read.csv(url("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv"), header = TRUE)
owid
dim(owid)
head(owid)
owid.lines
str(owid)
summary(owid)
```

###Full Cleaned Data Set
Data set, including possible confounders for use in later analysis, is cleaned and filtered below. 
```{r, eval = TRUE}
library(dplyr)
owid.filtered <- owid %>%
                  rename(iso.code = iso_code, totcases = total_cases, totdeaths = total_deaths, totcases.mil = total_cases_per_million, totdeaths.mil = total_deaths_per_million, hosp.patients = hosp_patients, hosp.patients.mil = hosp_patients_per_million, med.age = median_age, older70 = aged_70_older, gdp.cap = gdp_per_capita, card.death = cardiovasc_death_rate, diab.prev = diabetes_prevalence, fem.smokers = female_smokers, male.smokers = male_smokers) %>%
                  select(c(iso.code:totcases, totdeaths, totcases.mil, totdeaths.mil, hosp.patients, hosp.patients.mil, med.age, older70, gdp.cap, card.death, diab.prev, fem.smokers, male.smokers))
dim(owid.filtered)
head(owid.filtered)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
