---
title: "COVID-19 Outcomes"
author: "Alana Schreibman"
date: "5/27/2021"
output: 
  html_document:
    toc: true
    toc_float: 
        collapsed: true
        smooth_scroll: true
    depth: 3 
    theme: cosmo
    highlight: pygments
---

***

The purpose of this file is to practice applying basic R commands learned through lecture videos and perform exploratory analysis on worldwide COVID-19 data. 


### Our World in Data GitHub COVID-19 outcomes
Data was read in and summarized as shown below: (current file is as of 5/26)
```{r, eval = FALSE}
owid <- read.csv(url("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv"), na.strings = c("", "NA"), header = TRUE) 
dim(owid)
```

```{r, eval = FALSE}
head(owid)
str(owid)
summary(owid)
```

### Full Cleaned Data Set
Data set, including possible confounders for use in later analysis, is cleaned and filtered below. 
```{r, eval = FALSE}
library(dplyr)
owid.filtered <- owid %>%
                  rename(iso.code = iso_code, totcases = total_cases, totdeaths = total_deaths, totcases.mil = total_cases_per_million, totdeaths.mil = total_deaths_per_million, hosp.patients = hosp_patients, hosp.patients.mil = hosp_patients_per_million, med.age = median_age, older70 = aged_70_older, gdp.cap = gdp_per_capita, card.death = cardiovasc_death_rate, diab.prev = diabetes_prevalence, fem.smokers = female_smokers, male.smokers = male_smokers) %>%
                  select(c(iso.code:totcases, totdeaths, totcases.mil, totdeaths.mil, hosp.patients, hosp.patients.mil, med.age, older70, gdp.cap, card.death, diab.prev, fem.smokers, male.smokers))
dim(owid.filtered)
head(owid.filtered)
tail(owid.filtered)
table(owid.filtered$continent, useNA = "ifany") #Data includes both continent and country-specific outcomes. Continent outcomes can be distinguished by "NA" in the continent column; the continent name is in the location column. 
```

#### Continent Analysis
Data is filtered to remove entries for specific country locations and include only data that aggregates COVID-19 outcomes on a continent and world scale. 
```{r, eval = FALSE}
owid.continents <- owid.filtered %>%
                    filter(is.na(continent)) %>%
                    select(-continent)
print(owid.continents)
table(owid.continents$location)
```

#### Country Analysis
Data is filtered to include only country-specific COVID-19 outcomes and exclude continent and world-wide data.  
```{r, eval = FALSE}
owid.countries <- owid.filtered %>%
                    filter(!is.na(continent))
print(owid.countries)
table(owid.countries$continent)
```

#### Tests for Missingness
```{r, eval = TRUE}
t <- table(owid.filtered$totcases)
tail(t)
t.na <- table(owid.filtered$totcases, useNA = "ifany")
tail(t.na) #Keep this missing data in mind when analyzing case counts.  
```

#### Modified Data Frames
```{r, eval = FALSE}
#By continent:
owid.cases.cty <- select(owid.continents, c(iso.code:totcases, totcases.mil))
owid.deaths.cty <- select(owid.continents, c(iso.code:date, totdeaths, totdeaths.mil))
owid.hosp.cty <- select(owid.continents, c(iso.code:date, hosp.patients, hosp.patients.mil))

#By country:
owid.cases.cont <- select(owid.countries, c(iso.code:totcases, totcases.mil))
owid.deaths.cont <- select(owid.countries, c(iso.code:date, totdeaths, totdeaths.mil))
owid.hosp.cont <- select(owid.countries, c(iso.code:date, hosp.patients, hosp.patients.mil))
```

### Continent COVID-19 Plots
```{r, eval = TRUE}

```


countries:
group_by
owid.afghanistan <- owid.filtered %>%
                      select(c(location, date, totcases))
                      filter(location <- "Afghanistan")
print(owid.afghanistan)